\section{Contracts on interfaces}

\begin{frame}[t,fragile]{Repeating a contract}
\begin{itemize}
  \item Any redeclaration of a function has either the same contract or
completely omits the contract.
\end{itemize}
\vfill\pause
\begin{lstlisting}
int f(int x)
  [[expects: x>0]]
  [[ensures r: r >0]];

int f (int x) ; // OK. No contract.

int f ( int x)
  [[expects: x>=0]]; // Error missing ensures and different expects

int f(int x)
  [[expects: x>0]]
  [[ensures r: r >0]]; //OK. Same contract.
\end{lstlisting}
\end{frame}

\begin{frame}[t,fragile]{Preconditions on functions}
\begin{itemize}
  \item The expression of a precondition from a function may use:
    \begin{itemize}
      \item The function’s arguments.
      \item Any non-local object. 
    \end{itemize}
\end{itemize}
\begin{lstlisting}
constexpr int max = 100;
std::string name{"Daniel"};

bool f(int x, std::string s)
  [[expects: x>0]] // OK. x is an argument.
  [[expects: x<max]] // OK max is non-local
  [[expects: s.length()>0]] // OK. s is an argument
  [[expects: s!=name]]; // OK. name is non-local
\end{lstlisting}
\end{frame}

\begin{frame}[t,fragile]{Preconditions on \textbf{constexpr} functions}
\begin{itemize}
  \item The expression of a precondition from a \cppkey{constexpr} function may use:
    \begin{itemize}
      \item The function’s arguments.
      \item Any non-local object that is \cppkey{constexpr}.
      \item \textbad{but it cannot access non-local objects that are not
\cppkey{constexpr}}.
    \end{itemize}
\end{itemize}
\begin{lstlisting}
constexpr int max = 100;
std::string name{"Daniel"};

constexpr bool f(int x, std::string s)
  [[expects: x>0]] // OK. x is an argument.
  [[expects: x<max]] // OK max is constexpr
  [[expects: s.length()>0]] // OK. s is an argument
  [[expects: s!=name]]; // Error name is a non-local variable
\end{lstlisting}
\end{frame}

\begin{frame}[t,fragile]{Modifications in contracts}
\begin{itemize}
  \item A program with a contract expression that performs an
\textmark{observable modification} of an object is \textbad{ill-formed}.
    \begin{itemize}
      \item Your compiler might give a diagnostic.
    \end{itemize}
\end{itemize}

\vfill\pause
\begin{lstlisting}
int f(int x)
  [[expects: x++ > 0]]     // Error
  [[ensures r: r == ++x]]; // Error
\end{lstlisting}
\end{frame}

\begin{frame}[t,fragile]{Modified arguments and postconditions}
\begin{itemize}
  \item If a postcondition uses an argument and the function body modifies that
value, the program is ill-formed.
\end{itemize}

\vfill\pause
\begin{lstlisting}
int f(int x)
  [[ensures r: r==x]
{
  return ++x; // Error x used in postcondition
}
\end{lstlisting}
\pause
\begin{itemize}
  \item Workaround:
\end{itemize}
\begin{lstlisting}
int f(int x) {
  int oldx = x;
  auto r = ++x; 
  [[assert: r==oldx]];
}
\end{lstlisting}
\end{frame}

\begin{frame}[t,fragile]{But you con modify pointer contents}
\begin{itemize}
  \item A pointer value is different from the pointed value.
\end{itemize}

\vfill\pause
\begin{lstlisting}
void f(int * ptr)
  [[ensures: ptr!=nullptr]]
{
  *ptr = 42
}
\end{lstlisting}
\end{frame}

\begin{frame}[t,fragile]{Contracts in templated function}
\begin{itemize}
  \item The expression of a contract from a function template
        or a member function of a class template may use the
        template arguments.
\end{itemize}

\vfill\pause
\begin{lstlisting}
template <typename T, int size>
class table {
public:
  //...
  T & operator[](int i)
    [[expects: 0<=i && i<size]];
};
\end{lstlisting}

\end{frame}

\begin{frame}[t,fragile]{Contracts and visibility}
\begin{itemize}
  \item The contract from a public function shall not use members from protected
or private interfaces.
  \item The contract from a protected function shall not use members from private 
interface.
\end{itemize}

\vfill\pause
\begin{lstlisting}
template <typename T>
class table {
public:
  //...
  T & operator[](int i)
    [[expects: 0<=i && i<size_]]; // Error. size_ is private

private:
  //...
  int size_;
};
\end{lstlisting}
\end{frame}

\begin{frame}[t,fragile]{Contracts and function pointers}
\begin{itemize}
  \item A function pointer shall not include a contract.
  \item A call through a function pointer to functions with
a contract shall perform contract assertions checking once.
\end{itemize}

\vfill\pause
\begin{lstlisting}
using fpt = int (*)(int x)
  [[expects: x>=0]]
  [[ensures r: r>0]]; // Error. 

int g(int x) [[expects: x>=0]] [[ensures r: r>0]]
{
 return x+1;
}

int (*pf)(int) = g; // OK
\end{lstlisting}
\end{frame}

\begin{frame}[t,fragile]{Contracts and inheritance}
\begin{itemize}
  \item An \textmark{overriding function} shall have exactly the \textgood{same
contract} that was declared for that function in the base class.
    \begin{itemize}
      \item But the contract may be omitted in the overridden function.
    \end{itemize}
\end{itemize}

\vfill\pause
\begin{lstlisting}
struct B {
public:
  virtual void f (int x) [[expects: x>0]];
  //...
};

struct D : public B {
public:
  virtual void f (int x) override; // OK. expects: x>0
  //...
};
\end{lstlisting}
\end{frame}

\begin{frame}[t,fragile]{Contracts and inheritance}
\begin{itemize}
  \item An \textmark{overriding function} shall have exactly the \textgood{same
contract} that was declared for that function in the base class.
    \begin{itemize}
      \item Or it may be repeated.
    \end{itemize}
\end{itemize}

\vfill\pause

\begin{lstlisting}
struct B {
public:
  virtual void f (int x) [[expects: x>0]];
  //...
};

struct D : public B {
public:
  virtual void f (int x) override [[expects: x>0]]; // OK
  //...
};
\end{lstlisting}
\end{frame}

\begin{frame}[t,fragile]{Contracts and inheritance}
\begin{itemize}
  \item An \textmark{overriding function} shall have exactly the \textgood{same
contract} that was declared for that function in the base class.
    \begin{itemize}
      \item But the contract cannot be changed.
    \end{itemize}
\end{itemize}

\vfill\pause

\begin{lstlisting}
struct B {
public:
  virtual void f (int x) [[expects: x>0]];
  //...
};

struct D : public B {
public:
  virtual void f (int x) override [[expects: x!=0]]; // Error
  //...
};
\end{lstlisting}
\end{frame}

\begin{frame}[t,fragile]{Contracts and inheritance}
\begin{itemize}
  \item An \textmark{overriding function} shall have exactly the \textgood{same
contract} that was declared for that function in the base class.
    \begin{itemize}
      \item And it cannot be added.
    \end{itemize}
\end{itemize}

\vfill\pause

\begin{lstlisting}
struct B {
public:
  virtual void f (int x);
  //...
};

struct D : public B {
public:
  virtual void f (int x) override [[expects: x>0]]; // Error.
  //...
};
\end{lstlisting}
\end{frame}

\begin{frame}[t,fragile]{Precondition weakening}
\begin{itemize}
  \item Precondition weakening is not supported.
    \begin{itemize}
      \item But can be simulated.
    \end{itemize}
\end{itemize}

\vfill\pause
\begin{columns}[T]

\column{.5\textwidth}
\begin{lstlisting}
class A {
pubic:
  //...
  virtual void f(int x)
    [[expects: x>0]]
  {
    [[assert: x<max]];
    //..
  }
};
\end{lstlisting}

\column{.5\textwidth}
\begin{lstlisting}
class B : public A {
pubic:
  //...
  virtual void f(int x) override
    [[expects: x>0]]
  {
    //...
  }
};
\end{lstlisting}

\end{columns}
\end{frame}

\begin{frame}[t,fragile]{Postcondition strengthening}
\begin{itemize}
  \item Postcondition strengthening is not supported.
    \begin{itemize}
      \item but can be simulated.
    \end{itemize}
\end{itemize}

\vfill\pause
\begin{columns}[T]

\column{.5\textwidth}
\begin{lstlisting}
class A {
pubic:
  //...
  virtual int g()
    [[ensures r: r>=0]]
  {
    //..
  }
};
\end{lstlisting}

\column{.5\textwidth}
\begin{lstlisting}
class B : public A {
pubic:
  //...
  virtual int g() override
    [[ensures r: r>=0]]
  {
    //...
    [[assert: result<max]];
    return result;
  }
};
\end{lstlisting}

\end{columns}

\end{frame}
