\section{A brief history of contracts}

\begin{frame}[t,fragile]{What is contracts programming about?}
\begin{itemize}
  \item Improve how we write correct software.
    \begin{itemize}
      \item State preconditions, postconditions and assertions.
    \end{itemize}

  \vfill\pause
  \item Isn't a library solution enough?
    \pause
    \begin{itemize}
      \item We already tried that!
      \item Compilers and static analyzers do not understand that approach.
    \end{itemize}

  \vfill\pause
  \item What did others do?
    \pause
    \begin{itemize}
      \item Several language solutions out there (D, Ada, C\#, Microsoft SAL, \ldots).
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[t]{Contracts in C++}
\begin{itemize}
  \item First proposal for contracts programming in 2005.
    \begin{itemize}
      \item \textmark{N1613}: Proposal to add Design by Contract to C++.
            Throsten Ottosen.
      \item Died during the C++0x process.
    \end{itemize}

  \vfill\pause
  \item Next attempt in 2013.
    \begin{itemize}
      \item \textmark{N3604}: Centralized Defensive-Programming Support for
Narrow Contracts.
            John Lakos, Alexei Zakharov.
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[t]{Current contracts effort}
\begin{itemize}
  \item 2014: Multiple proposals on contracts programming.
    \begin{itemize}
      \item Discussions in the standards committee.
    \end{itemize}

  \vfill\pause
  \item 2016: Joint proposal trying to consider trade-offs.
    \begin{itemize}
      \item Gabriel Dos Reis, J. Daniel Garcia, John Lakos, Alisdair Meredith,
Nathan Myers, Bjarne Stroustrup.
    \vfill
      \item Targeting C++20.
    \end{itemize}

  \vfill\pause
  \item 2018: Approval.
    \begin{itemize}
      \item Part of the draft ISO/IEC 14882:2020.
    \end{itemize}
\end{itemize}
\end{frame}
